Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2015-09-06 pkg/glmertree/paper/treatment_interactions.Rnw
adapted R code to latest glmertree version -> still need to think about offset in empirical example
revised GLM-related descriptions somewhat



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@778 255f4249-2263-4417-b1ea-8498779e27a7

<<example-trees>>=
exampleglm <- lmtree(outcome ~ treatment | age, data = example)
exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data = example)
exampleglmmtree <- lmertree(outcome ~ treatment | age + duration + anxiety, outcome ~ (1|cluster), data = example)
@

\subsection{GLM}
In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM may be used to estimate treatment effects. The goal is to estimate a model for predicting the value of a treatment outcome, which follows, for example, a normal, binomial or Poisson distribution \footnote{ an overview of notation used is provided in the Appendix}:

\begin{equation}
\label{eq:expected_value}
	E[y_i | x_i] = \mu_i 
\end{equation}

\begin{equation}
\label{eq:fixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta 
\end{equation}

Where $y_i$ is the value of response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor $x_{i}^{\top}\beta$ and the mean of the response distribution function. In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}^2$. Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type). $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.

@@

@@

\begin{figure}[!th]
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
@
\caption{Example of a normal GLM (with fixed effects only) for treatment outcomes, based on the artificial motivating dataset ($N=150$). The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on the GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$th ($j = 1,\dots,J$) terminal node of the tree. As a result, in the GLM tree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':

\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}

Alternatively, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM. The model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{*}$ would also have length $2J$, and contain the subgroup-specific fixed-effects coefficients.

@@

\begin{figure}[!h]
\setkeys{Gin}{width=\textwidth}
<<exampleglmtree, fig=TRUE, width=12, height=9>>=
plot(exampleglmtree)
@
\caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
\label{fig:example_mobtree}
\end{figure}

\subsection{GLMM}
When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:

\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}

Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.

For simplicity, we assume that only cluster-specific intercepts are included in the models. However, random-effects covariates and coefficients can easily be included. Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the GLM.

@@

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMM tree, which combines the GLMM from Equation~\ref{eq:mixedeffects} with the tree from Equation~\ref{eq:fixedeffects_MOB}:

\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}

To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12}. They developed a method for the estimation of mixed-effects regression trees (MERTs), which are somewhat similar to GLMM trees. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the regression tree, and (2) assuming the regression tree known, allowing for estimation of the random effects.

@@

\vspace{5pt}
\noindent Step 0: Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.

\vspace{5pt}
\noindent \hangindent=4em Step 1: Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$), with $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.

\vspace{5pt}
\noindent Step 2: Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$, with $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.

\vspace{5pt}
\noindent Step 3: Repeat Steps~1 and~2 until convergence.

\vspace{5pt}
The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated. The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:mixedeffects}.

\begin{figure}[!h]
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

@@

\begin{figure}[!h]
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
  partynode(1L,
    split = partysplit(2L, breaks = 30),
    kids = list(
      partynode(2L,
        split = partysplit(1L, breaks = 17),
        kids = list(
          partynode(3L, info = c(
	    expression(''),
	    expression(beta[j0] == '17.5'),
	    expression(''),
	    expression(beta[j1] == -'5.0'),
	    expression(''),
	    expression(d[j] == -'1.0')
	  )),
          partynode(4L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )))),
      partynode(5L,
        split = partysplit(1L, breaks = 63),
        kids = list(
          partynode(6L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )),
          partynode(7L, info = c(
	    expression(''),
	    expression(beta[j0] == '42.5'),
	    expression(''),
	    expression(beta[j1] == '5.0'),
	    expression(''),
	    expression(d[j] == '1.0')
	  )))))),
  data.frame(U1 = numeric(0), U2 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

\begin{table}[!thbp]
	\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
	\small
	\begin{tabular}{lrrrrrrr}
		\thickline
		&\multicolumn{7}{c}{tree size}\\
		\cline{2-7}
		&   1    &   3   &   5   &   7   &   9   &  11  &   total \\
		\hline
		GLMM tree	& 20625 & 932 & 43 &0&0&0& 21,600\\
			  		& (.96) & (.04) & ($<.01$) &(.00)&(.00)&(.00)&(1.00) \\
   		GLM tree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
    			& (.67)  & (.19) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
		\hline
  		\multicolumn{8}{l}{\textit{Note. }Bracketed values are proportions. Tree sizes are expressed as the total}\\
  		\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $J$ nodes has $(J+1)/2$ }\\
  		\multicolumn{8}{l}{terminal nodes; the true tree size in datasets without treatment-subgroup}\\	  	\multicolumn{8}{l}{interactions was 1.}\\
	\end{tabular}
	\label{tab:treesize_main}
\end{table}

\begin{figure}[!thbp]
<<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
load("treesizes_long_nointeract.dat")
treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered = TRUE)
treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered = TRUE)
aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
	abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
	lines=TRUE, points=TRUE)))
@
\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=1$ represents the true tree size.}
\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

\begin{table}[!thbp]
\caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
\small
	\begin{tabular}{lrrrrrrrr}
		\thickline
		&&\multicolumn{7}{c}{tree size}\\
		\cline{2-8}
		&     3   &   5   &   7   &   9   &  11  &   13   &   15  & total \\
		\hline
	  	GLMM tree	& 2 & 281& 29358 & 2675 & 83 & 1 &0& 32,400 \\
					&  ($<.01$)  & ($<.01$)& (.91) & ($.08$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\

@@

\begin{figure}[!thbp]
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
	abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=7$ represents true tree size.}
\label{fig:xyplot_treesize_interact}
\end{figure}

@@

\begin{figure}[!thbp]
<<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.acc <- aggregate(formula=truetree.values ~ truetree.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=truetree.ind, type="b", 
	ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=TRUE, points=TRUE))
)
@
\caption{Average accuracy of GLM and GLMM trees. Accuracy of trees is defined as the proportion of datasets in which the true tree was accurately recovered. Values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

\begin{figure}[!thbp]
<<correlation-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, 
	type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average predictive accuracy of GLM and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2. Values $5$ and $2.5$ refer to the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences; values $200$, $500$ and $1000$ refer to sample size.}
\label{fig:xyplot_correlations}
\end{figure}

@@

To illustrate the application of, and differences in the results of GLM tree and GLMM tree, we applied both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}. This meta-analysis was based on individual-patient data from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression. The study of \citeA{CuijyWeit14} was aimed at establishing whether gender is a predictor or moderator of the outcomes of psychological and pharmacological treatments for depression. Treatment outcomes were assessed by means of the 17-item Hamilton Rating Scale for Depression (HAM-D; \citeNP{Hami60}). \citeA{CuijyWeit14} found no indication that gender either predicted or moderated treatment outcomes. Further details on the dataset are provided in \citeA{CuijyWeit14}.

In our analyses, posttreatment HAM-D score was the outcome variable, and potential partitioning variables were age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pretreatment HAM-D score. The predictor variable in the linear model was treatment type (0=CBT and 1=PHA). An indicator for study was used as the cluster indicator.

@@

We build all trees using data of patients with complete observations; that is, observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score. As a result, data from 694 patients from 7 studies were included for the analyses. Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \cite{CuijyWeit14}.

@@

<<depression-trees>>=
formula <- HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0

## GLM tree
lmtree_app <- lmtree(formula, data = metadata,
  offset = fitted(lm(HRSDt1 ~ HRSDt0), data = metadata))

## GLMM tree
lmertree_app <- lmertree(formula, HRSDt1 ~ (1 | studyid) + HRSDt0, data = metadata,
  ranefstart = fitted(lm(HRSDt1 ~ HRSDt0, data = metadata)))
@

>>>>

<<example-trees>>=
exampleglm <- lmtree(outcome ~ treatment | age, data = example)
exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data = example)
exampleglmmtree <- lmertree(outcome ~ treatment | cluster | age + duration + anxiety, data = example)
@

\subsection{GLM}
In a clinical trial, where the outcomes of two or more treatments are compared, an overall GLM is often used to estimate treatment effects. GLMs allow for the choice of a suitable response distribution -- for example normal, binomial, or Poisson - depending on whether the treatment outcome variable is continuous (e.g., an improvement score), binary (e.g., improved or not), or a count (e.g., number of events in a certain time span), respectively. In all cases the expectation $\mu_i$ of the outcome variable $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and a suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function\footnote{An overview of notation used is provided in the appendix.}. Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type). $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups. In case of a continuous response variable, we employ a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$.

@@

\begin{figure}[t!]
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
@
\caption{Example of a normal GLM (with fixed effects only) for treatment outcomes, based on the artificial motivating dataset ($N=150$). The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

@@

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on the GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node of the tree. As a result, in the GLM tree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':
%
\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}
%
Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all coefficients interact with the factor indicating the subgroup. Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{*}$ would also have length $2J$, and contain the subgroup-specific fixed-effects coefficients.

\begin{figure}[t!]
\setkeys{Gin}{width=\textwidth}
<<exampleglmtree, fig=TRUE, width=12, height=9>>=
plot(exampleglmtree)
@
\caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
\label{fig:example_mobtree}
\end{figure}

@@

\subsection{GLMM}
When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.

For simplicity, we assume that only cluster-specific intercepts are included in the models. However, random-effects covariates and coefficients can easily be included.

Note that, if the random-effects coefficients were known, the model could be estimated by a simple GLM as in Equation \ref{eq:fixedeffects} where $z_{i}^{\top}b$ would only be added as an offset (i.e., a variable with a fixed coefficient of 1) to the linear predictor.

@@

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMM tree, which combines the GLMM from Equation~\ref{eq:mixedeffects} with the tree from Equation~\ref{eq:fixedeffects_MOB}: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12} but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the regression tree, and (2) assuming the regression tree known, allowing for estimation of the random effects.

\begin{figure}[t!]
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

@@

\begin{description}
  \item[Step 0:] Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.

  \item[Step 1:] Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$), with $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.

  \item[Step 2:] Estimate random effects in the mixed effects model $x_{i}^{\top}\hat{\beta}_{j(r)} + z_{i}^{\top}\hat{b}_{(r)}$ with subgroups $j(r)$ from the GLM tree.
  % Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$, with $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.

  \item[Step 3:] Repeat Steps~1 and~2 until convergence.
\end{description}

The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated. The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:mixedeffects}.

@@

\begin{figure}[t!]
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
  partynode(1L,
    split = partysplit(2L, breaks = 30),
    kids = list(
      partynode(2L,
        split = partysplit(1L, breaks = 17),
        kids = list(
          partynode(3L, info = c(
	    expression(''),
	    expression(beta[j0] == '17.5'),
	    expression(''),
	    expression(beta[j1] == -'5.0'),
	    expression(''),
	    expression(d[j] == -'1.0')
	  )),
          partynode(4L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )))),
      partynode(5L,
        split = partysplit(1L, breaks = 63),
        kids = list(
          partynode(6L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )),
          partynode(7L, info = c(
	    expression(''),
	    expression(beta[j0] == '42.5'),
	    expression(''),
	    expression(beta[j1] == '5.0'),
	    expression(''),
	    expression(d[j] == '1.0')
	  )))))),
  data.frame(U1 = numeric(0), U2 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

\begin{table}[t!]
	\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
	\small
	\begin{tabular}{lrrrrrrr}
		\thickline
		&\multicolumn{7}{c}{tree size}\\
		\cline{2-7}
		&   1    &   3   &   5   &   7   &   9   &  11  &   total \\
		\hline
		GLMM tree	& 20625 & 932 & 43 &0&0&0& 21,600\\
			  		& (.96) & (.04) & ($<.01$) &(.00)&(.00)&(.00)&(1.00) \\
   		GLM tree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
    			& (.67)  & (.19) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
		\hline
  		\multicolumn{8}{l}{\textit{Note. }Bracketed values are proportions. Tree sizes are expressed as the total}\\
  		\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $J$ nodes has $(J+1)/2$ }\\
  		\multicolumn{8}{l}{terminal nodes; the true tree size in datasets without treatment-subgroup}\\	  	\multicolumn{8}{l}{interactions was 1.}\\
	\end{tabular}
	\label{tab:treesize_main}
\end{table}

\begin{figure}[t!]
<<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
load("treesizes_long_nointeract.dat")
treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered = TRUE)
treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered = TRUE)
aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
	abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
	lines=TRUE, points=TRUE)))
@
\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=1$ represents the true tree size.}
\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

\begin{table}[t!]
\caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
\small
	\begin{tabular}{lrrrrrrrr}
		\thickline
		&&\multicolumn{7}{c}{tree size}\\
		\cline{2-8}
		&     3   &   5   &   7   &   9   &  11  &   13   &   15  & total \\
		\hline
	  	GLMM tree	& 2 & 281& 29358 & 2675 & 83 & 1 &0& 32,400 \\
					&  ($<.01$)  & ($<.01$)& (.91) & ($.08$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\

@@

\begin{figure}[t!]
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
	abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=7$ represents true tree size.}
\label{fig:xyplot_treesize_interact}
\end{figure}

@@

\begin{figure}[t!]
<<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.acc <- aggregate(formula=truetree.values ~ truetree.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=truetree.ind, type="b", 
	ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=TRUE, points=TRUE))
)
@
\caption{Average accuracy of GLM and GLMM trees. Accuracy of trees is defined as the proportion of datasets in which the true tree was accurately recovered. Values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

\begin{figure}[t!]
<<correlation-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, 
	type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average predictive accuracy of GLM and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2. Values $5$ and $2.5$ refer to the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences; values $200$, $500$ and $1000$ refer to sample size.}
\label{fig:xyplot_correlations}
\end{figure}

@@

To illustrate the application of, and differences in the results of GLM tree and GLMM tree, we apply both algorithms to a dataset from a meta-analytic study of \citeA{CuijyWeit14}. This meta-analysis was based on individual-patient data from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression. The study of \citeA{CuijyWeit14} was aimed at establishing whether gender is a predictor or moderator of the outcomes of psychological and pharmacological treatments for depression. Treatment outcomes were assessed by means of the 17-item Hamilton Rating Scale for Depression (HAM-D; \citeNP{Hami60}). \citeA{CuijyWeit14} found no indication that gender either predicted or moderated treatment outcomes. Further details on the dataset are provided in \citeA{CuijyWeit14}.

In our analyses, posttreatment HAM-D score is the outcome variable, and potential partitioning variables are age, gender, level of education, presence of a comorbid anxiety disorder at baseline, and pretreatment HAM-D score. The predictor variable in the linear model was treatment type (0 = CBT and 1 = PHA). An indicator for study was used as the cluster indicator.

@@

We build all trees using data of patients with complete observations; that is, observations with non-missing values for potential partitioning variables, and pre- and posttreatment HAM-D score. As a result, data from 694 patients from 7 studies were included for the analyses. Results of our analysis may therefore not be representative of the complete dataset of the meta-analysis by \citeA{CuijyWeit14}.

@@

<<depression-trees>>=
hrs <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))

## GLM tree (estimated via lmertree to determine offset iteratively)
## lmtree_app <- lmtree(HRSDt1 ~ Tx_group | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
##   data = metadata, offset = hrs)
lmtree_app <- lmertree(HRSDt1 ~ Tx_group | HRSDt0 | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
  data = metadata, ranefstart = hrs, joint = FALSE)

## GLMM tree (estimated with joint = FALSE because fixed effect HRSDt0 is included in random formula)
lmertree_app <- lmertree(HRSDt1 ~ Tx_group | (1 | studyid) + HRSDt0 | Age + Gender + education + ComorbidAnxietyDisorder + HRSDt0,
  data = metadata, ranefstart = hrs, joint = FALSE)
@
